manuscript_figures

### 2024-07-16 scripts/manuscript_figures.Rmd
make fish production plots and updated calculations

Production to biomass ratios (P:B) and habitat productivity index for each fish species
```{r}
# load fish data from 2016
PB <- read.csv("data/fish.data.clean.calc.csv")

@@

# add total biomass for each transect per species
PB.aggregated <- PB_filtered %>%
  group_by(YEAR, SITE, TRANSECT, SP_CODE) %>%
  summarise(tot.biomass = sum(biomass, na.rm = TRUE),
            .groups = 'drop')

@@

# Convert YEAR to factor (if it's numeric)
PB_filtered$YEAR <- factor(PB_filtered$YEAR)

# Calculate production by multiplying biomass_per_m2 by PB raatio
PB_filtered$Production <- PB_filtered$PB.ratio * PB_filtered$biomass_per_m2

# make commmon name column
PB_filtered <- PB_filtered %>%
  mutate(Species = case_when(
    SP_CODE == "BRTY" ~ "Atlantic menhaden",
    SP_CODE == "CEST" ~ "Black sea bass",
    SP_CODE == "MOSA" ~ "Striped bass",
    SP_CODE == "PADE" ~ "Summer flounder",
    SP_CODE == "STCH" ~ "Scup",
    SP_CODE == "TAAD" ~ "Cunner",
    SP_CODE == "TAON" ~ "Tautog",
    TRUE ~ SP_CODE  # This will keep the original code if it doesn't match any of the above
  ))

# Now you can use this new 'Species' column in your ggplot
production_plot <- ggplot(PB_filtered, aes(x = YEAR, y = Production, fill = Species)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year",  # X-axis label
       y = expression("Fish Production" ~ (g ~ m^-2 ~ yr^-1)),  # Y-axis label
       fill = "Species") +  # This should now correctly set the legend title
  scale_fill_viridis_d() +  # Use viridis colorblind-friendly palette
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),  # Rotate x-axis labels if needed
        axis.title.x = element_text(vjust = -0.5, size = 18),  # Adjust x-axis title position
        axis.title.y = element_text(vjust = 1.5, size = 18),
        axis.text.y = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18)) +
  scale_y_continuous(limits = c(0, 4), expand = c(0, 0));production_plot  # Additional attempt to change legend title  # Use viridis colorblind-friendly palette

# Calculate Habitat Productivity Index (HPI), habitat productivity index (HPI), a proposed measure of habitat suitability and productive capacity, can be calculated as the product of average fish biomass and the species P/B coefficient (HPI = B × P/B) (Randall and Minns 2000)
# Calculate HPI
HPI_calculated <- PB_filtered %>%
  group_by(YEAR, SITE, TRANSECT) %>%
  summarise(HPI = sum(biomass_per_m2 * PB.ratio, na.rm = TRUE),
            .groups = 'drop')

@@

# Define the color palette
#my_palette <- brewer.pal(n = 6, name = "Paired")  # Adjust 'n' as per the number of years in your data

@@

data:  HPI by SITE
W = 527, p-value = 0.03417
alternative hypothesis: true location shift is not equal to 0

# Create box plot x axis as type (control and treatment) and y is HPI for kelp 
```{r}
# First, need to designate control and kelp treatments in the data frame for each year at each site, first filter 2016-2018 years because no 5 and 6 transects where used
# Filter out 2016-2018 YEAR data
HPI.kelp <- subset(morph.filt, YEAR < 2016 | YEAR > 2018)

>>>>

Production to biomass ratios (P:B) and fish produciton plots
```{r}
# load fish data from 2016
PB <- read.csv("data/fish.data.clean.calc.csv")

@@

# add total biomass for each transect per species
PB.aggregated <- PB_filtered %>%
  group_by(YEAR, SITE, TRANSECT, SP_CODE, PB.ratio, growth.coef) %>%
  summarise(tot.biomass = sum(biomass, na.rm = TRUE),
            .groups = 'drop')

@@

# Convert YEAR to factor (if it's numeric)
PB_filtered$YEAR <- factor(PB_filtered$YEAR)
PB.aggregated$YEAR <- factor(PB.aggregated$YEAR)

# Calculate production by methods of Ricker (1971) where P = G x B (P: production; G: instantaneous growth coefficient; B: average biomass)
# Step 1: Calculate average biomass
calc.prod <- PB.aggregated %>%
  group_by(YEAR, SITE, TRANSECT, SP_CODE, PB.ratio, growth.coef) %>%
  summarize(avg_biomass = mean(biomass_per_m2))

# Step 2: Calculate production
calc.prod$production <- calc.prod$growth.coef * calc.prod$avg_biomass

# make commmon name column
calc.prod <- calc.prod %>%
  mutate(Species = case_when(
    SP_CODE == "BRTY" ~ "Atlantic menhaden",
    SP_CODE == "CEST" ~ "Black sea bass",
    SP_CODE == "MOSA" ~ "Striped bass",
    SP_CODE == "PADE" ~ "Summer flounder",
    SP_CODE == "STCH" ~ "Scup",
    SP_CODE == "TAAD" ~ "Cunner",
    SP_CODE == "TAON" ~ "Tautog",
    TRUE ~ SP_CODE  # This will keep the original code if it doesn't match any of the above
  ))

# Now create a treatment column that assigns TRANSECTs 5 and 6 from each year as TREATMENT = control and all other transectes TREATMENT = kelp
calc.prod <- calc.prod %>%
  mutate(TREATMENT = ifelse(TRANSECT %in% c(5, 6), "control", "kelp"))

# create summary data for per year and treatment
pb_summary <- calc.prod  %>%
  group_by(YEAR, TREATMENT) %>%
  summarize(mean_production = mean(production, na.rm = TRUE),
            se_production = sd(production, na.rm = TRUE) / sqrt(n()))

# Now you can use this new 'Species' column in your ggplot
production_plot <- ggplot(pb_summary, aes(x = YEAR, y = mean_production, color = TREATMENT, group = TREATMENT)) +
  geom_point(size = 2) +         # Add points
  geom_line() +          # Add lines
  geom_errorbar(aes(ymin = mean_production - se_production, ymax = mean_production + se_production), width = 0.2) +  # Add error bars
  labs(x = "Year",  # X-axis label
       y = expression("Fish Production" ~ (g ~ m^-2 ~ yr^-1)),  # Y-axis label
       color = "Habitat") +  # Legend title
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),  # Rotate x-axis labels if needed
        axis.title.x = element_text(vjust = -0.5, size = 18),  # Adjust x-axis title position
        axis.title.y = element_text(vjust = 1.5, size = 18),
        axis.text.y = element_text(size = 18),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18)) +
  scale_y_continuous(limits = c(0, 8), expand = c(0, 0));production_plot # Additional attempt to change legend title  # Use viridis colorblind-friendly palette

# Save the plot
ggsave("output/figures/production.habitat.png", production_plot_A, width = 20, height = 10, units = "in", dpi = 300)

# Subset the data for control and kelp treatments
control_data <- subset(pb_summary, TREATMENT == "control")
kelp_data <- subset(pb_summary, TREATMENT == "kelp")

# Merge to align control and kelp data by year
merged_data <- merge(control_data, kelp_data, by = "YEAR", suffixes = c("_control", "_kelp"))

# Calculate the differences
merged_data$production_difference <- merged_data$mean_production_kelp - merged_data$mean_production_control
merged_data$se_difference <- sqrt(merged_data$se_production_kelp^2 + merged_data$se_production_control^2)

# Step 4: Plot the data
production_diff_plot <- ggplot(merged_data, aes(x = YEAR, y = production_difference, group = TREATMENT_kelp)) +
  geom_point(size = 2) +  # Points with light blue outline
  geom_line() +  # Lines with light blue color
  geom_errorbar(aes(ymin = production_difference - se_difference, ymax = production_difference + se_difference), 
                width = 0.2, color = "black", size = 0.7) +  # Error bars
  labs(x = "Year",  # X-axis label
       y = expression("Kelp Habitat Fish Production" ~ (g ~ m^-2 ~ yr^-1)),  # Y-axis label
       color = "Treatment") +  # Legend title
  theme_bw() +  # White background with black lines
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 18),  # Rotate x-axis labels if needed
        axis.title.x = element_text(vjust = -0.5, size = 18),  # Adjust x-axis title position
        axis.title.y = element_text(vjust = 1.5, size = 18),  # Adjust y-axis title position
        axis.text.y = element_text(size = 18),  # Adjust y-axis text size
        legend.text = element_text(size = 16),  # Legend text size
        legend.title = element_text(size = 18));production_diff_plot

# Save the plot
ggsave("output/figures/kelp.fish.production.png", production_diff_plot, width = 20, height = 10, units = "in", dpi = 300)

# Calculate the average of mean_production_kelp
average_production_kelp <- mean(merged_data$mean_production_kelp)

# Print the average
average_production_kelp # 4.359285 g/m2/yr
```

# Plot all two panels together using patchwork
```{r}
# Combine the plots with patchwork
prod_combined_plot <- (production_plot | production_diff_plot) +
  plot_annotation(tag_levels = 'A') + 
  theme(plot.tag = element_text(size = 20, face = "bold"));prod_combined_plot

# Calculate Habitat Productivity Index (HPI), habitat productivity index (HPI), a proposed measure of habitat suitability and productive capacity, can be calculated as the product of average fish biomass and the species P/B coefficient (HPI = B × P/B) (Randall and Minns 2000)
```{r}
# Calculate HPI
HPI_calculated <- PB_filtered %>%
  group_by(YEAR, SITE, TRANSECT) %>%
  summarise(HPI = sum(biomass_per_m2 * PB.ratio, na.rm = TRUE),
            .groups = 'drop')

@@

@@

data:  HPI by SITE
W = 527, p-value = 0.03417
alternative hypothesis: true location shift is not equal to 0
# Create box plot x axis as type (control and treatment) and y is HPI for kelp 
```{r}
# First, need to designate control and kelp treatments in the data frame for each year at each site, first filter 2016-2018 years because no 5 and 6 transects where used
# Filter out 2016-2018 YEAR data
HPI.kelp <- subset(morph.filt, YEAR < 2016 | YEAR > 2018)
